---
title: "STJ2012 analysis"
author: "Ross Cunning"
date: "July 28, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 4, digits = 2)
library(phyloseq); library(vegan)

taxcolors <- matrix(c(rgb(141, 211, 199, maxColorValue=255), rgb(255, 255, 179, maxColorValue=255), 
                      rgb(190, 186, 218, maxColorValue=255), rgb(251, 128, 114, maxColorValue=255), 
                      rgb(253, 180, 98, maxColorValue=255)), 
                    dimnames=list(c("CladeA", "CladeB", "CladeC", "CladeD", "CladeF")))
```

## Import data

```{r data_import, cache=T}
# Import OTU table
OTU <- otu_table(read.table("data/otu_table.tsv", header=T, check.names=F, row.names=1,
                            skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
# Import taxonomy
TAX <- tax_table(as.matrix(read.table("data/tax_assignments.txt", sep="\t", row.names=2,
                                      col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
# Import sample data
SAM <- sample_data(read.delim("data/mapping_file.txt", sep="\t", header=T, row.names=1))
# Build phyloseq object
phy <- phyloseq(OTU, TAX, SAM)
phy
```

## Descriptive statistics for dataset  
* Total count in OTU table: `r sum(otu_table(phy))`  
* Number of samples: `r nsamples(phy)`  
* Number of OTUs: `r ntaxa(phy)`  

#### Distribution of OTU counts and reads per sample  
```{r phy_histograms, fig.height=3, fig.width=8}
taxhist <- hist(log10(taxa_sums(phy)), plot=F)
samhist <- hist(log10(sample_sums(phy)), plot=F)

par(mfrow=c(1, 2), mar=c(3,3,1,1))
plot(taxhist, col="black", main=NA, xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(samhist, col="black", main=NA, xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
```

* Range of OTU counts: `r range(taxa_sums(phy))`  
* Number of singletons:  `r length(taxa_sums(phy)[taxa_sums(phy) <= 1])`  
* Range of reads per sample: `r range(sample_sums(phy))`  
* Mean (±SD) reads per sample (arithmetic): `r mean(sample_sums(phy))` ± `r sd(sample_sums(phy))`  
* Mean (±SD) reads per sample (geometric): `r exp(mean(log(sample_sums(phy))))` ± `r exp(sd(log(sample_sums(phy))))`  

```{r reads_per_sample, fig.height=2, fig.width=8}
plot_bar(phy, x="Sample", y="Abundance")
```

## Filter dataset  
#### Filter OTUs by minimum count
```{r filter_taxa}
# Set threshold count
n <- 10
# Identify OTUs below threshold count
taxa <- taxa_sums(phy)[which(taxa_sums(phy) >= n)]  
# Remove taxa below threshold count
phy.f <- prune_taxa(names(taxa), phy)  
```

* Number of OTUs remaining in dataset: `r ntaxa(phy.f)`

#### Filter samples by minimum count
```{r filter_samples}
# Set threshold number of reads
sn <- 10
# Remove samples with fewer reads than threshold
phy.f <- prune_samples(sample_sums(phy.f)>=sn, phy.f)
 
```

* Number of samples remaining in dataset: `r nsamples(phy.f) `

#### Distribution of OTU counts and reads per sample post-filtering
```{r phy.f_histograms, fig.height=3, fig.width=8}
taxhist <- hist(log10(taxa_sums(phy.f)), plot=F)
samhist <- hist(log10(sample_sums(phy.f)), plot=F)

par(mfrow=c(1, 2), mar=c(3,3,1,1))
plot(taxhist, col="black", main=NA, xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(samhist, col="black", main=NA, xlim=c(0, 5), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
```

* Range of OTU counts: `r range(taxa_sums(phy.f))`  
* Number of singletons:  `r length(taxa_sums(phy.f)[taxa_sums(phy.f) <= 1])`  
* Range of reads per sample: `r range(sample_sums(phy.f))`  
* Mean (±SD) reads per sample (arithmetic): `r mean(sample_sums(phy.f))` ± `r sd(sample_sums(phy.f))`  
* Mean (±SD) reads per sample (geometric): `r exp(mean(log(sample_sums(phy.f))))` ± `r exp(sd(log(sample_sums(phy.f))))`  

```{r reads_per_sample_filtered, fig.height=2, fig.width=8}
plot_bar(phy.f, x="Sample", y="Abundance")
```

## Transform taxa count data
```{r transform}
# Convert to proportion (relative abundance)
phy.f.p <- transform_sample_counts(phy.f, function(x) x/sum(x))
# Apply transformation function
transform <- function(x) sqrt(x/sum(x))  # Set transformation function
phy.f.t <- transform_sample_counts(phy.f, transform)  # Transform data
```

## Community composition in all samples
Here the composition of each sample is plotted as a horizontal bar, sorted by species and shore. Each segment of the bar represents an individual OTU. In the lefthand plot, all OTUs are colored by clade, in the right OTU, all OTUs are given a unique color.
```{r cladecomp, fig.height=6, fig.width=10}
# Define function to plot Symbiodinium community composition
composition <- function(phy, col, legend=T) {
  samdat <- data.frame(sample_data(phy))
  samdat$Genus <- factor(samdat$Genus, levels=rev(levels(samdat$Genus)))
  samdat$Species <- factor(samdat$Species, levels=rev(levels(samdat$Species)))
  samdat$Shore <- factor(samdat$Shore, levels=rev(levels(samdat$Shore)))
  samdat <- samdat[with(samdat, order(Genus, Species, Shore)), ]
  typerelabund <- as.matrix(otu_table(phy)[order(data.frame(tax_table(phy))$Subtype), rownames(samdat)])
  shorebreaks <- c(as.character(samdat$Shore), "X")==c("X", as.character(samdat$Shore))
  shorebreaks <- which(shorebreaks==F) - 1
  spbreaks <- c(which(duplicated(samdat$Species)==F) - 1, nrow(samdat))
  # Make Barplot
  barplot(typerelabund, horiz=T, space=0, axes=F,axisnames=F, yaxs="i", col=col)
  rect(0, 0, par("usr")[2], par("usr")[4], lwd=1, xpd=T)
  axis(side=1, at=seq(0, 1, 0.1), line=0, tck=-0.025, mgp=c(0,0.25,0), cex.axis=0.7)
  mtext(side=1, "Relative abundance", cex=0.7, line=1)
  # Add legend
  if (legend==T) {
    legend(x=par("usr")[2]/2, y=par("usr")[4], xjust=0.5, yjust=0.25, horiz=T, bty="n", xpd=T, cex=0.7, legend=c("A", "B", "C", "D", "F"), fill=taxcolors, x.intersp=0.5)
    legend(x=par("usr")[2]-0.03, y=par("usr")[4], xjust=0, yjust=0.1, bty="n", xpd=T, cex=0.6, pt.cex=0.6, legend=c("N Shore", "S Shore"), fill=c(0, "gray40"), y.intersp=0.7, x.intersp=0.3)
  }
  # Add grouping bars for Shore
  for (i in 1:length(shorebreaks)) {
    lines(c(0, 1), c(shorebreaks[i], shorebreaks[i]), lty=2, lwd=0.25)
    rect(1.01, shorebreaks[i], 1.04, shorebreaks[i+1], col=rep(c("gray40", 0), 50)[i], lwd=0.25, xpd=T)
  }
  # Add lines to separate species and species names
  for (i in 1:length(spbreaks)) {
    lines(c(0, 1.07), c(spbreaks[i], spbreaks[i]), xpd=T, type="l", lwd=0.4)
    text(1.03, (spbreaks[i] + spbreaks[i+1]) / 2, xpd=T, pos=4, cex=0.6,
        labels=paste(samdat$Genus[which(duplicated(samdat$Species)==F)][i], "\n",
                    samdat$Species[which(duplicated(samdat$Species)==F)][i], sep=""))
  }
}

par(mfrow=c(1,2), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7)
composition(phy.f.p, col=taxcolors[data.frame(tax_table(phy.f.p))[order(data.frame(tax_table(phy.f.p))$Subtype), ]$Clade], legend=T)
composition(phy.f.p, col=rainbow(ntaxa(phy.f.p)), legend=F)

```

# Statistical comparison of beta diversity defined by multivariate dispersion of B-C dissimilarity
#   Calculate distances to group centroids in full-dimensional principal coordinate space
#   Implemented by vegan betadisper(), based on Anderson (2006).
# Compartments/species:
```{r beta, cache=T}
sambd <- betadisper(d=vegdist(t(otu_table(phy.f.t)), method="bray"), 
                    group=data.frame(sample_data(phy.f.t))$Species, type="centroid")
anova(sambd)
boxplot(sambd)
sambdmeans <- aggregate(sambd$distances, by=list(group=sambd$group), FUN=mean)
sambdses <- aggregate(sambd$distances, by=list(group=sambd$group), 
                      FUN=function(x) sd(x)/sqrt(length(x)))
sampt <- permutest(sambd, pairwise = T)
sampt2 <- TukeyHSD(sambd)
library(multcompView)
saml <- multcompLetters(sampt$pairwise$permuted)#, Letters=c("e","a", "f","d","c","b"))
saml2 <- multcompLetters(sampt2$group[, 4])


# Figure: Distances to centroids
# Coral species
par(mfrow=c(1,1), mar=c(3.5,4,1,1), lwd=1)
plot(sambdmeans$x[order(sambdmeans$x, decreasing=T)], type="n", ylim=c(0, 0.75),
     ylab="Distance to centroid", xaxt="n", xlab="")
arrows(1:10, sambdmeans$x[order(sambdmeans$x, decreasing=T)] - sambdses$x[order(sambdmeans$x, decreasing=T)],
       1:10, sambdmeans$x[order(sambdmeans$x, decreasing=T)] + sambdses$x[order(sambdmeans$x, decreasing=T)],
       length=0.05, angle=90, code=3)
points(1:10, sambdmeans$x[order(sambdmeans$x, decreasing=T)], cex=1, pch=21, bg="white")
text(1:10, par("usr")[3]-0.025, srt=45, adj=1, xpd=T, cex=0.75,
     labels=levels(SAM$Species)[order(sambdmeans$x, decreasing=T)])

text(1:10, sambdmeans$x[order(sambdmeans$x, decreasing=T)] + sambdses$x[order(sambdmeans$x, decreasing=T)] + 0.03,
     labels=saml$Letters[as.character(sambdmeans$group)][order(sambdmeans$x, decreasing=T)], cex=0.5)
text(1:10, sambdmeans$x[order(sambdmeans$x, decreasing=T)] - sambdses$x[order(sambdmeans$x, decreasing=T)] - 0.03,
     labels=saml2$Letters[as.character(sambdmeans$group)][order(sambdmeans$x, decreasing=T)], cex=0.5)
     
```

```{r strigosa}
dstr <- subset_samples(phy.f.p, Species=="strigosa")
plot_bar(dstr, x="Sample", y="Abundance", fill="OTU")
plot_bar(dstr, x="Sample", y="Abundance", fill="Subtype")
# D strigosa hosted B1, B24, B35, B38, and C3 in Finney et al.... B25, B35, and B38 are missing from ref database.
# Also try 100% OTUs
```

     